\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}

### 2024-12-17 main.tex
tool basics implementated

\section{TODO}
\begin{itemize}
    \item Alpha equivalence is checked by a bound variable renaming on the whole term.
    \item arrow type of $\Type \to \Type$ are forbidden. Check whether typing rules comply with this restriction.
\end{itemize}

@@

\section{Language Syntax}
\begin{definition}[syntax]
    The syntax for type indices are defined as
    \begin{align*}
        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
    \end{align*}
    The syntax for Dirac notation types is defined as 
    \begin{align*}
        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall .T\ |\ \SET(\sigma). \\
    \end{align*}
    The syntax for Dirac notation terms is defined as
    \begin{align*}
        e ::=\ & x\ |\ \$i\ |\ (e_1, e_2)\ |\ \lambda : T.e\ |\ \lambda.e\ |\ e_1\ e_2 \\
        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
    \end{align*}
\end{definition}
 % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent. For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types. To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i : T . X$.

@@

\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & []\ |\ E; x : \Index\ |\ E; T : \Type\ |\ E; x : T\ |\ E; x := t : T. \\
        \Gamma ::=\ & []\ |\ \Gamma; \Index\ |\ \Gamma; T.
    \end{align*}
\end{definition}

Note that in the following rules, $x : \Index$ and $x : \Type$ are not considered as $x : T$ judgements. The context $\Gamma$ acts as the stack to indicate the typing of de Bruijn notations. The notation $\Gamma_i$ is defined as the $i$-th element in $\Gamma$ from the end. That is, if $\Gamma \equiv \Gamma'; T$, then $\Gamma_0 = T$ and $\Gamma_i = \Gamma'_{i-1}$.

\begin{figure}[h]
    \begin{gather*}
        \textbf{W-Empty} \qquad
        \frac{}{\WF([])[]} \\
        \\
        \textbf{W-AssumE-Index} \qquad
        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}\qquad \qquad
        \textbf{W-AssumE-Type} \qquad
        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Type)} \\
        \\
        \textbf{W-AssumE-Term} \qquad
        \frac{E[] \vdash T : \Type \qquad x \notin E}{\WF(E; x:T)[]}
        \qquad \qquad
        \textbf{W-Def-Term} \qquad
        \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]} \\
        \\
        \textbf{W-AssumC-Index} \qquad
        \frac{\WF(E)[\Gamma]}{\WF(E)[\Gamma; \Index]}
        \qquad \qquad
        \textbf{W-AssumC-Term} \qquad
        \frac{E[\Gamma] \vdash T : \Type}{\WF(E)[\Gamma; T]}
    \end{gather*}
    \caption{Rules for a well-formed environment and context.}
\end{figure}

\begin{figure}
    \begin{gather*}
        \textbf{Index-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E}{E[\Gamma] \vdash x : \Index}
        \qquad \qquad
        \textbf{Index-Prod} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index}
    \end{gather*}
    \caption{Rules for type index.}
\end{figure}

\begin{figure}
    \begin{gather*}
        \textbf{Type-Context-Var} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \$i : \Gamma_i}
        \\
        \\
        \textbf{Type-Arrow} \qquad
        \frac{E[\Gamma] \vdash T : \Type \qquad E[\Gamma] \vdash U : \Type}{E[\Gamma] \vdash T \to U : \Type}
        \qquad \qquad
        \textbf{Type-Index} \qquad
        \frac{E[\Gamma ; \Index] \vdash U : \Type}{E[\Gamma] \vdash \forall.U : \Type} \\
        \\        
        \textbf{Type-Basis} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \Basis(\sigma) : \Type}  \\
        \\
        \textbf{Type-Ket} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
        \qquad \qquad
        \textbf{Type-Bra} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \BType(\sigma) : \Type} \\
        \\
        \textbf{Type-Opt} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
        \qquad \qquad
        \textbf{Type-Scalar} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \SType : \Type} \\
        \\
        \textbf{Type-Set} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \SET(\sigma) : \Type}
    \end{gather*}
    \caption{Rules for types.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Term-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad (x:T) \in E \text{ or } (x := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
        \\
        \textbf{Lam} \qquad
        \frac{E[\Gamma; T] \vdash t : U}{E[\Gamma] \vdash (\lambda : T . t) : T \to U}
        \qquad \qquad
        \textbf{Index} \qquad
        \frac{E[\Gamma; \Index] \vdash t : U}{E[\Gamma] \vdash (\lambda.t) : \forall .U} \\
        \\
        \textbf{App-Arrow} \qquad
        \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T}
        \qquad \qquad
        \textbf{App-Index} \qquad
        \frac{E[\Gamma] \vdash t:\forall .U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U[u]}
    \end{gather*}
    \caption{Rules for variable and function typings. Here $U[u]$ means instantiate the first bound variable with $u$ in $U$.}
\end{figure}

@@

\clearpage
\section{Conversions and Reductions}
\begin{figure}[h]
    \begin{gather*}
        \textbf{Beta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda : T.t)\ u)\ \reduce\ t[u]} 
        \qquad \qquad
        \textbf{Beta-Index} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda.t)\ u)\ \reduce\ t[u]} 
        \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
        \\
        \textbf{Eta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ x)\ \reduce\ t}
        \qquad \qquad
        \textbf{Eta-Index} \qquad
        \frac{}{E[\Gamma] \vdash \lambda . (t\ x)\ \reduce\ t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

@@

\begin{align*}
    cmd ::=&\ \texttt{Def}(\texttt{ID}\ term) \\
        &|\ \texttt{Def}(\texttt{ID}\ term\ type) \\
        &|\ \texttt{Var}(\texttt{ID}\ term) \\
        &|\ \texttt{Check}(term) \\
        &|\ \texttt{Show}(\texttt{ID}) \\
        &|\ \texttt{ShowAll} \\
        &|\ \texttt{Normalize}(term) \\
        &|\ \texttt{CheckEq}(term\ term) \\
    type ::=&\ \texttt{Type}\ |\ \texttt{Arrow}(type\ type) \\
            &|\ \texttt{Base} \\
    term ::=&\ \texttt{Type}\ |\ \texttt{fun}(\texttt{ID}\ type\ term)\ |\ \texttt{apply}(term\ term)\ |\ \texttt{ID} \\
\end{align*}

>>>>

\section{Consideration}
\begin{itemize}
    \item AC equivalence and alpha equivalence.
    \item Efficient AC equivalence checking by sorting on syntax.
    \item Sorting requires that alpha equivalent terms have the same syntax.
    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
    \item only compute de Bruijn form in the equivalence checking and sorting phase.
\end{itemize}

\section{TODO}
\begin{itemize}
    \item implemented trace output for the whole pipeline.
    \item Note that the context should also be maintained during rewriting matching.
    \item better parser and output.
\end{itemize}

@@

\section{Language Syntax}
\begin{definition}[syntax]
    The syntax for type indices are defined as
    \begin{align*}
        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
    \end{align*}
    The syntax for Dirac notation types is defined as 
    \begin{align*}
        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
    \end{align*}
    The syntax for Dirac notation terms is defined as
    \begin{align*}
        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
    \end{align*}
\end{definition}
 % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent. For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types. To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{x \in s} \lambda x : T . X$ as well.

@@

\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & []\ |\ E; x : \Index\ |\ E; T : \Type\ |\ E; x : T\ |\ E; x := t : T. \\
        \Gamma ::=\ & []\ |\ \Gamma; x : \Index\ |\ \Gamma; x : T.
    \end{align*}
\end{definition}

Note that in the following rules, $x : \Index$ and $x : \Type$ are not considered as $x : T$ judgements.

\begin{figure}[h]
    \begin{gather*}
        \textbf{W-Empty} \qquad
        \frac{}{\WF([])[]} \\
        \\
        \textbf{W-AssumE-Index} \qquad
        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}\qquad \qquad
        \textbf{W-AssumE-Type} \qquad
        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Type)[]} \\
        \\
        \textbf{W-AssumE-Term} \qquad
        \frac{E[] \vdash T : \Type \qquad x \notin E}{\WF(E; x:T)[]}
        \qquad \qquad
        \textbf{W-Def-Term} \qquad
        \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]} \\
        \\
        \textbf{W-AssumC-Index} \qquad
        \frac{\WF(E)[\Gamma]}{\WF(E)[\Gamma; x : \Index]}
        \qquad \qquad
        \textbf{W-AssumC-Term} \qquad
        \frac{E[\Gamma] \vdash T : \Type}{\WF(E)[\Gamma; x : T]}
    \end{gather*}
    \caption{Rules for a well-formed environment and context.}
\end{figure}

\begin{figure}
    \begin{gather*}
        \textbf{Index-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
        \qquad \qquad
        \textbf{Index-Prod} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index}
    \end{gather*}
    \caption{Rules for type index.}
\end{figure}

\begin{figure}
    \begin{gather*}
        \textbf{Type-Arrow} \qquad
        \frac{E[\Gamma] \vdash T : \Type \qquad E[\Gamma] \vdash U : \Type}{E[\Gamma] \vdash T \to U : \Type}
        \qquad \qquad
        \textbf{Type-Index} \qquad
        \frac{E[\Gamma ; x : \Index] \vdash U : \Type}{E[\Gamma] \vdash \forall x.U : \Type} \\
        \\        
        \textbf{Type-Basis} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \Basis(\sigma) : \Type}  \\
        \\
        \textbf{Type-Ket} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
        \qquad \qquad
        \textbf{Type-Bra} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \BType(\sigma) : \Type} \\
        \\
        \textbf{Type-Opt} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
        \qquad \qquad
        \textbf{Type-Scalar} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \SType : \Type} \\
        \\
        \textbf{Type-Set} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \SET(\sigma) : \Type}
    \end{gather*}
    \caption{Rules for types.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Term-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad (x:T) \in E[\Gamma] \text{ or } (x := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
        \\
        \textbf{Lam} \qquad
        \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
        \qquad \qquad
        \textbf{Index} \qquad
        \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U} \\
        \\
        \textbf{App-Arrow} \qquad
        \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T}
        \qquad \qquad
        \textbf{App-Index} \qquad
        \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
    \end{gather*}
    \caption{Rules for variable and function typings. Here $U[u]$ means instantiate the first bound variable with $u$ in $U$.}
\end{figure}

@@

\clearpage
\section{Conversions and Reductions}

\begin{figure}[h]
    \begin{gather*}
        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}} \\
        \\
        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
        \\
        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
    \end{gather*}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Beta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda x : T.t)\ u)\ \reduce\ t\{x/u\}} 
        \qquad \qquad
        \textbf{Beta-Index} \qquad
        \frac{}{E[\Gamma] \vdash ((\mu x.t)\ u)\ \reduce\ t\{x/u\}} 
        \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
        \\
        \textbf{Eta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash \lambda x : T . (t\ x)\ \reduce\ t}
        \qquad \qquad
        \textbf{Eta-Index} \qquad
        \frac{}{E[\Gamma] \vdash \mu x . (t\ x)\ \reduce\ t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

@@

\begin{align*}
    cmd ::=&\ \texttt{Def}(\texttt{ID}\ term) \\
        &|\ \texttt{Def}(\texttt{ID}\ term\ type) \\
        &|\ \texttt{Var}(\texttt{ID}\ term) \\
        &|\ \texttt{Check}(term) \\
        &|\ \texttt{Show}(\texttt{ID}) \\
        &|\ \texttt{ShowAll} \\
        &|\ \texttt{Normalize}(term)\ |\ \texttt{Normalize}(term\ \textsf{Trace}) \\
        &|\ \texttt{CheckEq}(term\ term) \\
    type ::=&\ \texttt{Type}\ |\ \texttt{Arrow}(type\ type) \\
            &|\ \texttt{Base} \\
    term ::=&\ \texttt{Type}\ |\ \texttt{fun}(\texttt{ID}\ type\ term)\ |\ \texttt{apply}(term\ term)\ |\ \texttt{ID} \\
\end{align*}
